Correlations between gender and research topics at three major archaeology conferences

### 2020-11-12 analysis/paper/paper.Rmd
added mention of ostapenko2018textual

Related work on gender differences in research topics has noted major differences in the proportions of men and women across many academic-related interests. One widely discussed theory, especially in psychology, to explain this is the theory of people-thing interest dimensions which proposes that men have stronger interest in things and their mechanisms, while women have a stronger interest in persons and their feelings [@su2009men]. However, recent work by @thelwall2019gender shows that gender differences in choice of research topics cannot be fully explained by people-thing theory from psychology. Thewall et al. looked at 508,283 journal articles, classified the articles into research areas, and determined the authors' genders by matching names with the 1990 US census data. Instead of a people-thing contrast, they find that women more likely to use exploratory and qualitative methods rather than quantitative methods, and men show more interest in male in abstraction, and power/control fields. Similarly, ostapenko2018textual... In our data, we can similarly see in an interest from men in quantitative methods. However, our results do not so clearly divide the interests of the two genders we have data on, for example, with women are more prominent in GIS, and isotope analyses, which are highly quantitative areas.

>>>>

Related work on gender differences in research topics has noted major differences in the proportions of men and women across many academic-related interests. One widely discussed theory, especially in psychology, to explain this is the theory of people-thing interest dimensions which proposes that men have stronger interest in things and their mechanisms, while women have a stronger interest in persons and their feelings [@su2009men]. However, recent work by @thelwall2019gender shows that gender differences in choice of research topics cannot be fully explained by people-thing theory from psychology. Thewall et al. looked at 508,283 journal articles, classified the articles into research areas, and determined the authors' genders by matching names with the 1990 US census data. Instead of a people-thing contrast, they find that women more likely to use exploratory and qualitative methods rather than quantitative methods, and men show more interest in male in abstraction, and power/control fields.  In our data, we can similarly see in an interest from men in quantitative methods. However, our results do not so clearly divide the interests of the two genders we have data on, for example, with women are more prominent in GIS, and isotope analyses, which are highly quantitative areas.

@ostapenko2018textual identified common topics in the personal statements of aspiring surgeons that were specific to men and women authors. Women tended to discuss surgery as working as a team, while men focused on their specific individual clinical experiences. @ostapenko2018textual propose that the differences between male and female statements may come from actual motivating factors for career goals and aspirations, or they may reflect differences in beliefs about what makes a successful personal statement. They could not evaluate which explanation was most important, but a key observation is that authors of the personal statements may not deliberately choose the themes for their statements. Instead, they may be using themes that they have received positive feedback on on the past, that has perhaps unconsciously biased them towards certain themes in their writing. Similarly archaeologists might choose topics for their research based on a pattern of positive or negative feedback on topic choices over time, and so their choices are not fully deliberate.
